---
title: "Wastewater Metatranscriptomics: End-to-End Analysis"
subtitle: "Rothman et al. 2021 — Viral Surveillance in Wastewater Influent"
author: "Kelvin Wong"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6
)

library(dplyr)
library(ggplot2)
library(tidyr)
library(jsonlite)
library(knitr)
library(kableExtra)
library(plotly)

# All output files live in this directory
DATA_DIR <- "~/metagenomics_job"
```

---

## Study Overview

This notebook presents an end-to-end metatranscriptomic analysis of paired wastewater samples
(enriched vs. unenriched) collected from the Encina Water Pollution Control Facility in Carlsbad,
California. The data derive from **Rothman et al. 2021** (*Applied and Environmental Microbiology*),
which profiled the wastewater virome across seven treatment plants between August 2020 and January 2021.

**Ten samples** spanning five consecutive timepoints (Oct 5 – Nov 2, 2020) were selected: one
unenriched and one respiratory-panel-enriched replicate per date.

| Step | Tool | Purpose |
|------|------|---------|
| Download & downsample | `fasterq-dump`, `seqtk` | Retrieve ~10 M bp per sample |
| QC | `fastp` | Trim adapters, filter low-quality reads |
| Host filtering | `bowtie2` vs GRCh38 | Remove human reads |
| Taxonomy | `kraken2` + `bracken` | Classify reads against a viral DB; re-estimate species-level abundance |
| Variant calling | `bowtie2` + `bcftools` | Call SARS-CoV-2 SNVs in enriched samples |

---

## Steps 1 & 2: QC and Host Filtering

### Analysis

Raw paired-end reads were quality-trimmed with **fastp** (adapter removal, Q < 15 sliding window,
minimum length 30 bp). Cleaned reads were then aligned to the human reference genome (GRCh38)
using **bowtie2**; read pairs that failed to align were retained as host-depleted reads for
downstream analysis.

Host depletion is essential for wastewater metatranscriptomics: human nucleic acids shed from
gastrointestinal and epithelial cells can comprise the vast majority of sequencing reads, masking
the environmental viral signal.

### Implementation

```{r qc-load}
srr_ids <- c(
  "SRR18340978", "SRR18341017", "SRR18341028",
  "SRR18341039", "SRR18341080",
  "SRR19606889", "SRR19606932", "SRR19606954",
  "SRR19607206", "SRR19607228"
)

# Parse fastp JSON for each sample
parse_fastp <- function(srr) {
  path <- path.expand(file.path(DATA_DIR, paste0(srr, "_fastp.json")))
  js   <- fromJSON(path)
  tibble(
    Sample            = srr,
    Reads_Before_QC   = js$summary$before_filtering$total_reads,
    Reads_After_QC    = js$summary$after_filtering$total_reads,
    Q30_Before        = round(js$summary$before_filtering$q30_rate * 100, 1),
    Q30_After         = round(js$summary$after_filtering$q30_rate  * 100, 1),
    Pct_Reads_Kept    = round(js$summary$after_filtering$total_reads /
                                js$summary$before_filtering$total_reads * 100, 1)
  )
}

qc_stats <- bind_rows(lapply(srr_ids, parse_fastp))
```

```{r qc-host-counts}
# Count host-depleted reads from the output FASTQ files (read 1 only; pairs are equal)
count_reads <- function(srr) {
  f <- path.expand(file.path(DATA_DIR, paste0(srr, "_host_removed_1.fastq")))
  raw <- system(paste("wc -l <", shQuote(f)), intern = TRUE)
  n   <- as.integer(trimws(raw))
  n / 4L   # 4 lines per FASTQ record
}

host_removed_counts <- sapply(srr_ids, count_reads)

qc_stats <- qc_stats %>%
  mutate(
    Reads_After_HostFilter = host_removed_counts * 2L,   # both reads
    Pct_Host_Removed = round(
      (1 - host_removed_counts * 2 / Reads_After_QC) * 100, 1)
  )
```

```{r qc-metadata}
# Load sample metadata
metadata <- read.csv(path.expand(file.path(DATA_DIR, "SraRunTable.csv")),
                     stringsAsFactors = FALSE) %>%
  select(Run, Collection_Date, Library.Name) %>%
  mutate(
    Treatment = ifelse(
      grepl("enriched", Library.Name, ignore.case = TRUE),
      "Enriched", "Unenriched"
    )
  ) %>%
  rename(Sample = Run)

qc_full <- qc_stats %>%
  left_join(metadata, by = "Sample") %>%
  select(Sample, Collection_Date, Treatment,
         Reads_Before_QC, Reads_After_QC, Q30_After,
         Reads_After_HostFilter, Pct_Host_Removed)
```

```{r qc-table}
qc_full %>%
  arrange(Collection_Date, Treatment) %>%
  kable(
    col.names = c("Sample", "Date", "Treatment",
                  "Reads (raw)", "Reads (post-QC)", "Q30 % (post-QC)",
                  "Reads (host-filtered)", "% Host Removed"),
    caption = "Table 1. QC and host-depletion summary for all 10 samples.",
    format.args = list(big.mark = ",")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(which(qc_full %>% arrange(Collection_Date, Treatment) %>%
                   pull(Treatment) == "Enriched"),
           background = "#eaf4fb")
```

```{r qc-plot}
qc_plot_data <- qc_full %>%
  arrange(Collection_Date, Treatment) %>%
  mutate(Label = Collection_Date) %>%
  select(Label, Treatment,
         `Raw reads`     = Reads_Before_QC,
         `Post-QC`       = Reads_After_QC,
         `Host-depleted` = Reads_After_HostFilter) %>%
  pivot_longer(c(`Raw reads`, `Post-QC`, `Host-depleted`),
               names_to = "Stage", values_to = "Reads") %>%
  mutate(Stage = factor(Stage, levels = c("Raw reads", "Post-QC", "Host-depleted")))

p1 <- ggplot(qc_plot_data, aes(x = Label, y = Reads / 1e3, fill = Stage,
                                text = paste0("Sample: ", Label,
                                              "<br>Stage: ", Stage,
                                              "<br>Reads (k): ", round(Reads / 1e3, 1)))) +
  geom_col(position = "dodge") +
  facet_wrap(~ Treatment, scales = "free_x") +
  scale_fill_manual(values = c("#4e79a7", "#f28e2b", "#59a14f")) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Figure 1. Read counts at each processing stage",
    x     = NULL,
    y     = "Read pairs (thousands)",
    fill  = "Stage"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1, size = 9),
        legend.position = "top")

ggplotly(p1, tooltip = "text", width = 750) %>%
  layout(
    title  = list(text = "Figure 1. Read counts at each processing stage",
                  y = 0.93, x = 0, xanchor = "left"),
    legend = list(orientation = "v", x = 1.15, y = 0.5),
    xaxis  = list(tickfont = list(size = 8)),
    xaxis2 = list(tickfont = list(size = 8)),
    margin = list(t = 80)
  )
```

### Multiple Choice Question — QC

> **Quality Control**
>
> After running fastp on your wastewater metatranscriptomic samples, you observe that fewer than
> **1% of reads are discarded**, yet the mean read length decreases from **101 bp to ~90 bp**.
> Which of the following best explains this?
>
> **A)** The sequencing run failed to generate full-length reads, indicating a library preparation error.
>
> **B)** Most reads are high quality end-to-end, but soft-trimming of low-quality bases and adapter
> sequences at read ends shortens reads without discarding them entirely. ✓
>
> **C)** The fastp parameters were set too permissively, retaining reads that should have been removed.
>
> **D)** The reduction in read length indicates that the majority of reads are derived from degraded
> RNA fragments rather than intact viral transcripts.

### Multiple Choice Question — Host Filtering

> **Host Filtering**
>
> After running bowtie2 host depletion on your wastewater metatranscriptomic samples, you find
> that **fewer than 5% of reads** map to the human reference genome across all samples. Which of
> the following best explains this surprisingly low rate of host reads?
>
> **A)** The host depletion step failed — the reads were not properly filtered against the human genome.
>
> **B)** Wastewater is a complex environmental matrix where human nucleic acids are heavily diluted
> by microbial, dietary, and environmental genetic material, so low host-mapping rates are
> expected and do not indicate a failed depletion. ✓
>
> **C)** The GRCh38 reference genome is too incomplete to capture most human sequences in sewage.
>
> **D)** The samples were collected from a low-density residential area with minimal human
> biological input into the sewershed.

---

## Steps 3 & 4: Taxonomic Composition and Treatment Differences

### Analysis

Host-depleted reads were classified against a viral Kraken2 database using the
**lowest common ancestor (LCA)** algorithm. Species-level relative abundances were then
re-estimated with **Bracken**, which redistributes reads assigned to higher-level taxonomic
nodes (genus, family) back down to the species level using a Bayesian statistical model.
The resulting abundance profiles were compared between enriched and unenriched samples
across the five timepoints, with SARS-CoV-2 relative abundance tracked as a key indicator
of community transmission.

> **Note on Bracken re-estimation:**
> Kraken2 assigns reads using a lowest common ancestor (LCA) algorithm, which can leave a large
> fraction of reads attributed to higher-level taxonomic nodes (e.g., genus or family) rather than
> to a specific species. **Bracken** (Bayesian Re-estimation of Abundance with KrakEN) addresses
> this by redistributing those higher-level read counts down to the species level using a
> statistical model built from the database's k-mer distributions, producing more accurate
> species-level relative abundances — particularly for low-abundance taxa that Kraken2 tends to
> underestimate. The analysis below uses Bracken-corrected abundances (`fraction_total_reads`)
> with `-r 100` to match the post-QC mean read length of ~90 bp.

### Implementation

```{r classification-summary}
# Read classified vs unclassified counts from Kraken2 reports
kraken_files <- list.files(path.expand(DATA_DIR), pattern = "_kraken\\.report$",
                           full.names = TRUE)

classif_data <- lapply(kraken_files, function(file) {
  df <- read.table(file, sep = "\t", header = FALSE, stringsAsFactors = FALSE, quote = "")
  colnames(df) <- c("percentage", "count_frag", "count_taxon", "rank", "taxid", "name")
  df$name <- trimws(df$name)
  srr <- regmatches(basename(file), regexpr("SRR[0-9]+", basename(file)))
  unclassified <- df$count_frag[df$rank == "U"]
  classified   <- df$count_frag[df$rank == "R"]
  total        <- unclassified + classified
  tibble(
    Sample       = srr,
    Total_Reads  = total,
    Classified   = classified,
    Unclassified = unclassified,
    Pct_Classified = round(classified / total * 100, 1)
  )
}) %>%
  bind_rows() %>%
  left_join(metadata, by = "Sample") %>%
  arrange(Collection_Date, Treatment)
```

```{r classification-plot}
classif_plot_data <- classif_data %>%
  mutate(
    Label = Collection_Date,
    Unclassified_pct = Unclassified / Total_Reads * 100,
    Classified_pct   = Classified   / Total_Reads * 100
  ) %>%
  select(Label, Treatment, Collection_Date, Classified_pct, Unclassified_pct) %>%
  pivot_longer(c(Classified_pct, Unclassified_pct),
               names_to = "Category", values_to = "Percentage") %>%
  mutate(Category = factor(ifelse(Category == "Classified_pct",
                                  "Classified", "Unclassified"),
                           levels = c("Unclassified", "Classified")))

p_class <- ggplot(classif_plot_data,
                  aes(x = reorder(Label, Collection_Date), y = Percentage,
                      fill = Category,
                      text = paste0("Sample: ", Label,
                                    "<br>", Category, ": ",
                                    round(Percentage, 1), "%"))) +
  geom_col(position = "stack") +
  facet_wrap(~ Treatment, scales = "free_x") +
  scale_fill_manual(values = c("Classified" = "#59a14f", "Unclassified" = "#d3d3d3")) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Figure 2. Viral Classification Rate: Classified vs. Unclassified Reads",
    x     = NULL,
    y     = "Percentage of Host-Depleted Reads (%)",
    fill  = NULL
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1, size = 9),
        legend.position = "top")

ggplotly(p_class, tooltip = "text", width = 700) %>%
  layout(
    legend = list(orientation = "v", x = 1.15, y = 0.5),
    margin = list(t = 80)
  )
```

```{r taxonomy-load}
bracken_files <- list.files(path.expand(DATA_DIR), pattern = "_bracken_species\\.out$",
                            full.names = TRUE)

viral_data <- lapply(bracken_files, function(file) {
  df <- read.table(file, sep = "\t", header = TRUE,
                   stringsAsFactors = FALSE, quote = "")
  df$Sample <- regmatches(basename(file),
                          regexpr("SRR[0-9]+", basename(file)))
  df
}) %>%
  bind_rows() %>%
  inner_join(metadata, by = "Sample") %>%
  rename(Species = name, Relative_Abundance = fraction_total_reads)
```

```{r taxonomy-topspecies}
top_viruses <- viral_data %>%
  group_by(Species) %>%
  summarise(total = sum(Relative_Abundance), .groups = "drop") %>%
  slice_max(total, n = 12) %>%
  pull(Species)

plot_data <- viral_data %>%
  mutate(Plot_Species = ifelse(Species %in% top_viruses, Species, "Other")) %>%
  group_by(Sample, Collection_Date, Treatment, Plot_Species) %>%
  summarise(Abundance = sum(Relative_Abundance), .groups = "drop")
```

```{r taxonomy-plot}
# Pre-compute proportions so tooltip values match the filled bar heights
plot_data_norm <- plot_data %>%
  group_by(Collection_Date, Treatment) %>%
  mutate(Proportion = Abundance / sum(Abundance)) %>%
  ungroup()

p2 <- ggplot(plot_data_norm,
       aes(x = as.factor(Collection_Date), y = Proportion, fill = Plot_Species,
           text = paste0("Date: ", Collection_Date,
                         "<br>Species: ", Plot_Species,
                         "<br>Proportion: ", round(Proportion, 4)))) +
  geom_bar(stat = "identity", position = "stack",
           color = "white", linewidth = 0.2) +
  facet_wrap(~ Treatment) +
  scale_fill_viridis_d(option = "turbo") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Figure 3. Viral Taxonomic Composition: Enriched vs. Unenriched Wastewater",
    x     = "Collection Date",
    y     = "Relative Abundance (Proportion)",
    fill  = "Viral Species"
  ) +
  theme(
    axis.text.x        = element_text(angle = 45, hjust = 1),
    legend.position    = "right",
    legend.text        = element_text(size = 9),
    panel.grid.major.x = element_blank()
  )

ggplotly(p2, tooltip = "text") %>%
  layout(
    legend = list(orientation = "v", x = 1.02, y = 0.5),
    xaxis  = list(tickfont = list(size = 10), tickangle = -45),
    xaxis2 = list(tickfont = list(size = 10), tickangle = -45),
    margin = list(t = 80)
  )
```

```{r sars-abundance}
# Extract SARS-CoV-2 relative abundance per sample from Bracken output
sars_data <- viral_data %>%
  filter(grepl("severe acute", Species, ignore.case = TRUE)) %>%
  select(Sample, Collection_Date, Treatment, Relative_Abundance) %>%
  # Add zero rows for samples with no SARS-CoV-2 detected
  right_join(metadata %>% left_join(
    viral_data %>% select(Sample, Collection_Date) %>% distinct(),
    by = "Sample"), by = c("Sample", "Treatment")) %>%
  mutate(Relative_Abundance = replace_na(Relative_Abundance, 0),
         Collection_Date = coalesce(Collection_Date.x, Collection_Date.y)) %>%
  select(Sample, Collection_Date, Treatment, Relative_Abundance) %>%
  distinct()

p3 <- ggplot(sars_data,
             aes(x = as.Date(Collection_Date), y = Relative_Abundance * 100,
                 color = Treatment, group = Treatment,
                 text = paste0("Date: ", Collection_Date,
                               "<br>Treatment: ", Treatment,
                               "<br>SARS-CoV-2: ", round(Relative_Abundance * 100, 2), "%"))) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Enriched" = "#e15759", "Unenriched" = "#4e79a7")) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week") +
  theme_minimal(base_size = 12) +
  labs(
    title = "Figure 4. SARS-CoV-2 Relative Abundance Over Time",
    x     = "Collection Date",
    y     = "SARS-CoV-2 Relative Abundance (%)",
    color = "Treatment"
  )

ggplotly(p3, tooltip = "text") %>%
  layout(margin = list(t = 80))
```

### Multiple Choice Question — Step 3: Taxonomic Composition

> **Taxonomic Composition**
>
> Your Kraken2/Bracken classification shows that the **majority of host-depleted reads cannot
> be assigned to any known virus** across most samples — including 3 out of 5 enriched samples
> which show classification rates below 6%, comparable to unenriched samples. Only 2 enriched
> samples (Oct 12 and Oct 26) show substantially higher classification (~32%). Which of the
> following best explains why most reads remain unclassified regardless of treatment?
>
> **A)** The host-depletion step was too aggressive and inadvertently removed most viral reads
> along with the human reads.
>
> **B)** Unclassified reads in wastewater arise from two compounding factors: environmental
> viruses such as bacteriophages and plant-derived organisms that are absent from current
> reference databases, and RNA degradation in sewage that fragments and damages captured
> sequences enough to break Kraken2's k-mer matching — even for known target viruses like
> SARS-CoV-2. ✓
>
> **C)** The enrichment panel failed for most samples and the experiment needs to be repeated.
>
> **D)** A low classification rate indicates the sequencing runs were contaminated and the
> reads are mostly noise.

### Multiple Choice Question — Step 4: Treatment Differences

> **Treatment Differences**
>
> When comparing your enriched and unenriched wastewater sequencing results, you notice the
> unenriched samples are almost entirely dominated by **plant viruses**, with SARS-CoV-2
> completely undetected, while enriched samples show SARS-CoV-2 comprising **25–63%** of
> classified reads. Which of the following is best supported?
>
> **A)** The unenriched samples were heavily contaminated by agricultural runoff before reaching
> the treatment plant.
>
> **B)** The bioinformatic classification database is fundamentally biased against human viruses.
>
> **C)** Plant viruses from food consumption represent the true ecological baseline of the
> wastewater virome, and the Illumina Respiratory Virus Oligonucleotide Panel's
> hybridisation probes were necessary to selectively capture and concentrate the low-abundance
> SARS-CoV-2 signal above this environmental background. ✓
>
> **D)** The community had no active COVID-19 cases on the days the unenriched samples were collected.

---

## Step 5: SARS-CoV-2 Single Nucleotide Variants (SNVs)

### Analysis

Host-depleted reads from enriched samples were aligned to the Wuhan-Hu-1 SARS-CoV-2 reference
genome using **bowtie2** (local, very-sensitive). Variants were called with **bcftools mpileup /
call**. Because wastewater RNA is highly degraded, genomic coverage is uneven ("Swiss cheese"
pattern). To avoid reporting noise as signal, SNVs were only retained at positions with a minimum
total read depth of **≥ 10** reads and an alternate allele frequency of **≥ 5%**.

The interactive heatmap below shows variant frequencies across samples and genomic positions,
allowing rapid identification of mutations shared across timepoints — a signal of sustained
community transmission — versus private variants present in only one sample.

> Variant calling was run on all five enriched samples:
> SRR19606889, SRR19606932, SRR19606954, SRR19607206, SRR19607228.

### Implementation

```{r variants-load}
df_variants <- read.csv(path.expand(file.path(DATA_DIR, "variant_summary.csv")))

df_filtered <- df_variants %>%
  filter(!is.na(POS),
         Total_Depth >= 10,
         Frequency   >= 0.05) %>%
  arrange(POS) %>%
  left_join(metadata, by = "Sample")
```

```{r variants-summary-table}
df_filtered %>%
  count(Sample, Treatment, name = "N_SNVs") %>%
  arrange(Treatment, Sample) %>%
  kable(caption = "Table 3. Number of SNVs passing filters per sample.",
        col.names = c("Sample", "Treatment", "SNVs (depth ≥ 10, freq ≥ 5%)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r variants-heatmap, fig.height=8}
fig <- plot_ly(
  data       = df_filtered,
  x          = ~Sample,
  y          = ~as.factor(POS),
  z          = ~Frequency,
  type       = "heatmap",
  colorscale = "Viridis",
  hoverinfo  = "text",
  text       = ~paste(
    "Sample:", Sample,
    "<br>Treatment:", Treatment,
    "<br>Position:", POS,
    "<br>Mutation:", REF, "->", ALT,
    "<br>Depth:", Total_Depth,
    "<br>Frequency:", round(Frequency, 3)
  )
) %>%
  layout(
    title  = "Figure 5. SARS-CoV-2 Variant Frequencies Across Samples",
    xaxis  = list(title = "Sample ID"),
    yaxis  = list(title = "Genomic Position",
                  type  = "category",
                  tickmode = "auto",
                  nticks   = 20),
    margin = list(l = 100, b = 100, t = 80)
  )

fig
```

### Multiple Choice Question

> **SARS-CoV-2 SNV Analysis**
>
> You are tracking SARS-CoV-2 variants in a wastewater sample. For a specific
> lineage-defining mutation, your variant caller outputs an **alternate allele depth of 5** and a
> **total read depth of 5** (yielding a 100% mutation frequency). Which of the following is
> best supported?
>
> **A)** The variant has reached complete fixation and is the only strain circulating in the
> local community.
>
> **B)** While the calculated frequency is 100%, a total depth of 5 is statistically
> indistinguishable from sequencing noise in degraded wastewater and cannot be confidently
> reported. ✓
>
> **C)** The sample must be chemically re-enriched because the genomic coverage is uneven.
>
> **D)** The reference genome was incorrectly assigned during the alignment phase.

---

## Session Info

```{r session-info}
sessionInfo()
```
